Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-08-07 comments/paper_MMC_discussion.Rnw
some of aristotle's comments addressed

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))

@@

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable}
\usepackage{titlesec, siunitx}
\usepackage[round,authoryear]{natbib}
\usepackage[section]{placeins}  % keep floats in their place.
\usepackage{authblk}            % Listing Author affiliations
\usepackage[hypcap]{caption}

@@

% styling
\newcommand{\subsubsubsection}[1]{\paragraph{#1}}
\renewcommand\Affilfont{\itshape\small}  %authblk

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T }
\newcommand{\FEPdataset}{SZ-FEP }

% title and authors
\title{Bootstrapping Multi-atlas Hippocampal Segmentation with \mb}
\author[1]{Jon Pipitone}
\author[1]{Matt T. Park}
\author[1]{Julie Winterburn}
\author[1]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4]{Jens C. Pruessner}
\author[4]{Martin Lepage}
\author[1,5]{Aristotle Voineskos}
\author[1,5,6]{M. Mallar Chakravarty}
\author[ ]{Alzheimer's Disease Neuroimaging Initiative}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health, Hospital for Sick Children, Toronto,
ON,Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Brain Imaging Group, Douglas Hospital Research Centre, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[6]{Institute of Biomedical Engineering, University of Toronto, Toronto,
ON, Canada}

\renewcommand\Authands{ and }

\date{}

\begin{document}
\maketitle

\begin{abstract}
%Neuroimaging research often relies on automated anatomical segmentations of MR
%images of the brain. Current multi-atlas based approaches provide accurate
%segmentations of brain images by propagating manually derived segmentations of
%specific neuroanatomical structures to unlabelled data. These approaches often
%rely on a large number of such manually segmented atlases that take significant
%time and expertise to produce. We present an algorithm for the automatic
%segmentation of the hippocampus that minimizes the number of atlases needed
%while still achieving similar accuracy to multi-atlas approaches.
\end{abstract}

\parbox{4in}{
\textbf{Contact:} \\
Jon Pipitone and M. Mallar Chakravarty \\
Kimel Family Translation Imaging-Genetics Research Laboratory \\
Research Imaging Centre \\
Centre for Addiction and Mental Health \\
250 College St. \\
Toronto, Canada   M5T 1R8 \\
jon.pipitone@camh.ca; mallar.chakravarty@camh.ca}

@@

The hippocampus is a neuroanatomical structure situated in the medial temporal
lobe of the brain, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}.  In addition to its known functional roles,
the hippocampus is of interest to neuroscientists because it is implicated in
several forms of brain disfunction such as Alzheimer's disease
\citep{Sabuncu2011} and schizophrenia \citep{Narr2004,Karnik-Henry2012}.  In
neuroimaging experiments, magnetic resonance images (MRI) are often used for the
identification of the hippocampus.  As such, accurate segmentation of the
hippocampus and its subregions in MRI is a necessary first step to better
understand the unique neuroanatomy of subjects. Typically, the gold standard for
neuroanatomical segmentation is manual delineation by an expert human rater. However
the rapid increase in the availability of MRI data and the time and expertise
required for manual segmentation is prohibitive
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}.
Further, there is little agreement between researchers regarding how exactly the
hippocampus should be identified in MRI images \citep{Geuze2004} and this has
led to efforts to create an unified hippocampal segmentation protocol
\citep{Jack2011}.

Automated segmentation techniques that are reliable, objective, and reproducible
are a necessary alternative to manual segmentation.  In the case of classical
model-based segmentation methods \citep{Haller1997,Csernansky1998}, an MRI atlas
that was previously manually labelled by an expert rater is matched to target
images using nonlinear registration methods.  The resulting nonlinear
transformation is applied to the manual labels (i.e. {\em label propagation}) to
move them into the target image space.  While this methodology has been used
successfully in several contexts
\citep{Chakravarty2008,Chakravarty2009,Collins1995,Haller1997}, it is limited
in accuracy by the introduction of errors due inaccuracies in the nonlinear
transformation itself, partial volume effects in label resampling, and
irreconcilable differences between the neuroanatomy represented within the atlas
and target images.

One methodology that can be used to mitigate these sources of errors involves
the use of multiple manually segmented atlases and probabilistic segmentation
techniques, such as those found in the FreeSurfer package \citep{Fischl2002}.
FreeSurfer uses a probabilistic atlas of anatomical and tissue classes along
with spatial constraints for class labels encoded using a Markov random field
model to segment the entire brain.

More recently, many groups have been using multiple atlases to improve overall
segmentation accuracy (i.e. multi-atlas segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into a definitive segmentation for the target.  In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such a cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations.  For example \citet{Collins2010}
demonstrated that only 14 atlases, selected based on highest similarity between
medial temporal lobe neuroanatomy as evaluated by normalized mutual information
\citep{Studholme1999} from a library of 80 atlases, were required to achieve
accurate segmentations of the hippocampus.  Several methods have been explored
for label fusion including the STAPLE algorithm \citep{Warfield2004} that
computes a probabilistic segmentation using an expectation maximization
framework from an set of competing segmentations; or others where a subset of
segmentations can be estimated using metrics such as the sum of squared
differences in the regions of interest to be segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library; ranging from atlas libraries
that require between 30 \citep{Heckemann2006} and 80 \citep{Collins2010}
manually segmented atlases.  This strategy has the main drawback of being
inflexible as it does not easily accommodate varying the definition of the
hippocampal anatomy (such as the commonly used heuristic of subdividing the
hippocampus in to head, body, and tail \citep{Poppenk2011,Pruessner2000}).
Furthermore, none of these methods have demonstrated sufficient flexibility to
accommodate atlases that are somehow exceptional such as those derived from
serial histological data \citep{Chakravarty2006,Yelnik2007} or high-resolution MRI
data that enables robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}.  Due to
the recent availability of the latter, there has been increased interest in the
use of probabilistic methods for the identification of the hippocampal subfields
on standard T1-weighted images.  Our group recently demonstrated that through
use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{MallarChakravarty2012}.
The novelty of this manuscript is the extension our multi-atlas methodology to
the hippocampus using more than a single input atlas, while simultaneously
limiting the number of possible inputs used during segmentation, and
demonstrating that accurate identification of the hippocampal subfields is
indeed possible using this methodology.

There are few methods that have attempted to perform multi-atlas segmentation
with a limited number of input atlases. The LEAP algorithm is an elegant
modification to the basic multi-atlas strategy \citep{Wolz2010} in which the
atlas library is grown, beginning with a set of manually labelled atlases, and
successively incorporating unlabelled target images after themselves being
labelling using multi-atlas techniques. The sequence in which target images are
labelled is chosen so that the similarity between the atlas images and the
target images is minimised at each step, effectively allowing for deformations
between very dissimilar images to be broken up into sequences of smaller
deformations.  Although \citet{Wolz2010} begin with an atlas library of 30 MR
images, this method could theoretically work using a much smaller atlas library.
In their validation, LEAP was used to segment the whole hippocampus in the
ADNI-1 baseline dataset, achieving a mean Dice score of 0.85 with manual
segmentations.

To the best of our knowledge there are two other segmentation methods that
attempt to define the hippocampal subfields using standard T1-weighted data.
The first is included with the FreeSurfer package \citep{VanLeemput2009}.  This
work is limited as it omits the tail of the hippocampus and the segmentation
protocol has yet to be fully validated.  Nonetheless, they demonstrate that the
applicability of their work using data from 10 subjects.  In the second method,
\citet{Yushkevich2009} acquired and labelled hippocampal subfields on
high-resolution MRI data from post-mortem medial temporal lobe samples. Using
nonlinear registration guided by manually derived hippocampus masks and specific
landmarks, they demonstrate accurate parcellation of hippocampal subfields in
unlabelled MRI volumes.

@@

The simplest form of multi-atlas segmentation, (which we will call {\em basic 
multi-atlas segmentation}), involves three steps.  First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propagated to the target image space. 
Third, the labels are combined into a single labelling by way of a label fusion method 
\citep{Heckemann2006, Heckemann2011}. This method is described in detail in other
publications \citep{Collins2010,Heckemann2011,Aljabar2009}.

@@

The following section describes the experiments we conducted to assess the
segmentation quality of the \mb algorithm. The first two experiments assess the
validity of \mb using a cross-validation design. Experiment 1 investigates the
accuracy of \mb whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 tests hippocampal subfield segmentation quality. The last two
experiments assess the validity of the \mb algorithm when applied to different
diseases: first episode schizophrenia (Experiment 3), and Alzheimer's disease
(Experiment 4). Additionally, in experiment 4, we compare \mb segmentations with
those of well-known automated and manual methods and assess segmentation bias.

@@

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Manual segmentations of the left and
right whole hippocampi are available \citep{Hsu2002}. These labels have been
generated using the SNT tool from Medtronic Surgical Navigation Technologies,
Louisville, CO (see Supplementary Materials for detailed discussion of the
manual segmentation process used).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% ADNI1 X-Validation Dataset Demographics Table
%
<<ADNI-Xval-demographics, echo=F, results="asis",cache=FALSE>>=
adni_69_RIDs = c(295,413,619,685,729,782,938,954,1018,1155,907,981,222,324,448,
                 546,553,572,602,610,814,1341,675,681,731,1130,41,68,70,101,
                 249,293,316,414,698,1206,1222,1339,751,1030,3,5,10,16,22,53,
                 168,183,241,1205,328,1095,991,213,159,337,343,642,753,1109,
                 1217,29,56,1188,1293,149,382,431,1202)
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               adni_69_RIDs, method = "reverse", test=FALSE, overall=TRUE)
caption = paste(
  "\\textbf{ADNI-1 cross-validation subset demographics.}", 
  "Clinical Dementia Rating-Sum of Boxes (CDR-SB), Alzheimer's Disease",
  "Assessment Scale (ADAS), Mini–Mental State Examination (MMSE).", sep=" ") 
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      label="tab:ADNI1-xvalidation-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Monte Carlo Cross-Validation (MCCV), also known as repeated random sub-sampling
cross-validation, consists of repeated rounds of validation conducted on a fixed
dataset \citep{Shao1993}. In each round, the dataset is randomly partitioned into
a training set and a validation set. The method to be validated is then given
the training data, and its output is compared with the validation set.

@@

A total of ten validation rounds are performed on each subject in the dataset,
over each combination of parameter settings. The parameter settings
we explore are: atlas library size (1-9), template library size (1-20),
registration method (\ants or \animal), and label fusion method (majority vote,
cross-correlation weighted majority vote, and normalized mutual information
weighted majority vote).  A total of $10 \times 69 \times 9 \times 20 \times 2
\times 3 = \num{7452000}$ validation rounds were conducted, resulting in a total
of $\num{1490400}$ segmentations analysed.

@@

The \animal algorithm carries out image registration is in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the whole
brain \citep{Collins}. In the second phase, nonlinear registration is completed
using the \animal algorithm \citep{Collins1995}: an iterative procedure that
estimates a 3D deformation field between two MR images. At first, large
deformations are estimated using blurred version of the input data. These larger
deformations are then input to subsequent steps where the fit is refined by
estimating smaller deformations on data blurred with a Gaussian kernel with a
smaller FWHM. The final transformation is a set of local translations defined on
a bed of equally spaced nodes that were estimated through the optimization of
the correlation coefficient.  For the purposes of this work we used the
regularization parameters optimized in \citet{Robbins2004}, displayed in Table
\ref{tab:ANIMAL-params}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ANIMAL parameters                    
%                                     
\begin{table*}[!tbp]
\scriptsize
\caption{\textbf{\animal registration parameters.} \label{tab:ANIMAL-params}}
\begin{center}
\begin{tabular}{l | c c c}
\hline 
Parameters        & Stage 1 & Stage 2 & Stage 3  \tabularnewline
\hline 
Model Blur (FWHM) &   8     &    8    &   4      \tabularnewline
Input Blur (FWHM) &   8     &    8    &   4      \tabularnewline
Iterations        &   30    &    30   &   10     \tabularnewline
Step              &   8x8x8 &  4x4x4  &   2x2x2  \tabularnewline
Sub-Lattice       &   6     &    6    &   6      \tabularnewline
Lattice Diameter  &24x24x24 &12x12x12 & 6x6x6    \tabularnewline
\hline
\end{tabular}
\end{center}
\end{table*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\ants is a diffeomorphic registration algorithm which provides great flexibility
over the choice of transformation model, objective function, and the consistency
of the final transformation \citep{Avants2008}. The transformation is estimated in a
hierarchical fashion where the MRI data is subsampled, allowing large
deformations to be estimated and successively refined at later hierarchical
stages (where the data is subsampled to a finer grid). The deformation field and
the objective function are regularized with a Gaussian kernel at each level of
the hierarchy. The \ants algorithm is freely available
\url{http://www.picsl.upenn.edu/ANTS/}. We used an implementation of the \ants
algorithm compatible with the MINC data format, mincANTS
\url{https://github.com/vfonov/mincANTS}.

@@

The Dice similarity coefficient (DSC) assesses the agreement between two
segmentations. It is one of the most widely used measures of segmentation
performance, and we use it as the basis of comparison in this experiment.
%Additionally, we report the Jaccard index, another commonly used similarity
%measure:

@@

% \[\text{Jaccard (J)} = \frac{|A \cap B|}{|A \cup B|} = \frac{DSC}{(2-DSC)}\]

@@

The manual segmentations (SNT) provided as part of the ADNI dataset are used as
the gold standard to compare with \citep{Hsu2002}.  The segmentation accuracy
reported is averaged over the ten validation rounds for each parameter setting.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw())

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  require(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

bland_altman_plot <- function(x, y, c) {
  data = data.frame(x=x,y=y,diff=x-y, m = (x+y)/2, c=c)
  return(ggplot(data, aes(x= m, y = diff, colour = as.factor(c))) + 
    geom_hline(yintercept = mean(data$diff) + c(-2, 0, 2) * sd(data$diff), 
               linetype=2, color='brown') + geom_point())
}

@@

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable}
\usepackage{titlesec, siunitx}
\usepackage[round,authoryear]{natbib}
\usepackage[section]{placeins}  % keep floats in their place.

@@

% add \subsubsubsection
% \setcounter{secnumdepth}{4}  
% \titleformat{\paragraph}
% {\normalfont\normalsize\bfseries}{\theparagraph}{1em}{}
% \titlespacing*{\paragraph}
% {0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}
\newcommand{\subsubsubsection}[1]{\paragraph{#1}}

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T }
%notes and TODO formatting
%\newcommand{\marginnote}[1]{\-\marginpar[\raggedleft\footnotesize #1]{\raggedright\footnotesize #1}}
%\newcommand{\comment}[1]{\begin{kframe}{\textcolor{red}{#1}}\end{kframe}}
%\newcommand{\todo}[1]{\comment{TODO #1}}
%\newcommand{\mc}[1]{\comment{MC: #1}}
\begin{document}
%\SweaveOpts{concordance=TRUE}

\title{Bootstrapping Multi-atlas Hippocampal Segmentation with \mb}
\author{Pipitone J., Winterburn J., Lett, T., Lerch J., Pruessner J., Lepage M., \\ 
Voineskos A., Chakravarty M.M., and \\ 
the Alzheimer's Disease Neuroimaging Initiative}
\maketitle

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating manually derived segmentations of
specific neuroanatomical structures to unlabelled data. These approaches often
rely on a large number of such manually segmented atlases that take significant
time and expertise to produce. We present an algorithm for the automatic
segmentation of the hippocampus that minimizes the number of atlases needed
while still achieving similar accuracy to multi-atlas approaches.
\todo[inline]{finish}

\end{abstract}

@@

The hippocampus is a neuroanatomical structure situated in the medial temporal
lobe of the brain, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}.  In addition to its known functional roles,
the hippocampus is of interest to neuroscientists because it is implicated in
several forms of brain disfunction such as Alzheimer's disease
\citep{Sabuncu2011} and schizophrenia \citep{Narr2004,Karnik-Henry2012}.  In
neuroimaging experiments, magnetic resonance images (MRI) are often used for the
identification of the hippocampus.  As such, accurate segmentation of the
hippocampus and its subregions in MRI is a necessary first step to better
understand the unique neuroanatomy of subjects. Typically, the gold standard for
neuroanatomical segmentation is manual delineation by an expert human. However
the rapid increase in the availability of MRI data and the time and expertise
required for manual segmentation is prohibitive
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007,Weiner2013}.
Further, there is little agreement between researchers regarding how exactly the
hippocampus should be identified in MRI images \citep{Geuze2004} and this has
led to efforts to create an unified hippocampal segmentation protocol
\citep{Jack2011}.

Automated segmentation techniques that are reliable, objective, and reproducible
are a necessary alternative to manual segmentation.  In the case of classical
model-based segmentation methods \citep{Haller1997,Csernansky1998}, an MRI atlas
that was previously manually labelled by an expert rater is matched to target
images using nonlinear registration methods.  The resulting nonlinear
transformation is applied to the manual labels (ie. {\em label propagation}) to
apply them to the target image.  While this methodology has been used
successfully in several contexts
\citep{MallarMedIA200809,MallarHBM20082009,Collins1995,Haller1997}, it is limited
in accuracy by the introduction of errors due inaccuracies in the nonlinear
transformation itself, partial volume effects in label resampling, and
irreconcilable differences between the neuroanatomy represented within the atlas
and target images.

One methodology that can be used to mitigate these sources of errors involves
the use of multiple manually segmented atlases and probabilistic segmentation
techniques, such as those found in the FreeSurfer package \citep{Fischl2002},
FreeSurfer uses a probabilistic atlas of anatomical and tissue classes along
with spatial constraints for class labels encoded using a Markov random field
model to segment the entire brain.

More recently, many groups have been using multiple atlases to improve overall
segmentation accuracy (ie. multi-atlas segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into a definitive segmentation for the target.  In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such a cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations.  For example Collins and Pruessner \citep{Collins2010}
demonstrated that only 14 atlases, selected based on highest similarity between
medial temporal lobe neuroanatomy as evaluated by normalized mutual information
\citep{Studholme1999} from a library of 80 atlases, were required to achieve
accurate segmentations of the hippocampus.  Several methods have been explored
for label fusion including the STAPLE algorithm \citep{Warfield2004} that
computes a probabilistic segmentation using an expectation maximization
framework from an set of competing segmentations; or others where a subset of
segmentations can be estimated using metrics such as the sum of squared
differences in the regions of interest to be segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library; ranging from atlas libraries
that require between 30 \citep{Heckemann2006} and 80 \citep{Collins2010}
manually segmented atlases.  This strategy has the main drawback of being
inflexible as it does not easily accommodate varying the definition of the
hippocampal anatomy (such as the commonly used heuristic of subdividing the
hippocampus in to head, body, and tail \citep{Poppenk2011,Pruessner2000}).
Furthermore, none of these methods have demonstrated sufficient flexibility to
accommodate atlases that are somehow exceptional such as those derived from
serial histological data \citep{Chakravarty06a,Yelnik2007} or high-resolution MRI
data that enables robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}.  Due to
the recent availability of the latter, there has been increased interest in the
use of probabilistic methods for the identification of the hippocampal subfields
on standard T1-weighted images.  Our group recently demonstrated that through
use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data \citep{Chakravarty2013} is possible.  The
novelty of this manuscript is the extension our multi-atlas methodology to the
hippocampus using more than a single input atlas, while simultaneously limiting
the number of possible inputs used during segmentation, and demonstrating that
accurate identification of the hippocampal subfields is indeed possible using
this methodology.

There are few methods that have attempted to perform multi-atlas segmentation
with a limited number of input atlases. The LEAP algorithm is an elegant
modification to the basic multi-atlas strategy \citep{Wolz2010} in which the
atlas library is grown, beginning with a set of manually labelled atlases, and
successively incorporating unlabelled target images after themselves being
labelling using multi-atlas techniques. The sequence in which target images are
labelled is chosen so that the similarity between the atlas images and the
target images is minimised at each step, effectively allowing for deformations
between very dissimilar images to be broken up into sequences of smaller
deformations.  Although Wolz et al. begin with an atlas library of 30 MR images,
this method could theoretically work using a much smaller atlas library.  In
their validation, LEAP was used to segment the whole hippocampus in the ADNI1
baseline dataset, achieving a mean Dice score of 0.85 with manual segmentations.

To the best of our knowledge there are two other segmentation methods that
attempt to define the hippocampal subfields using standard T1-weighted data.
The first is included with the FreeSurfer package \citep{VanLeemput2009}.  This
work is limited as it omits the tail of the hippocampus and the segmentation
protocol has yet to be fully validated.  Nonetheless, they demonstrate that the
applicability of their work using data from 10 subjects.  In the second method,
Yushkevich and colleagues \citep{Yushkevich2009} acquired and labelled
hippocampal subfields on high-resolution MRI data from post-mortem medial
temporal lobe samples an demonstrate it's applicability in the segmentation of
other MRI-volumes using nonlinear registration guided using manually derived
hippocampus masks and specific landmarks. In their work they demonstrate
accurate parcellations of the subfields.

@@

The simplest form of multi-atlas segmentation, (which we will call {\em basic 
multi-atlas segmentation}), involves three steps.  First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propogated to the target image space. 
Third, the labels are combined into a single labelling by way of a label fusion method 
\citep{Heckemann2006, Heckemann2011}. This method is described in detail in other
publications \citep{Collins2010,Heckemann2011,Aljabar2009}.

@@

The following section describes the experiments we conducted to assess the
segmentation quality of the \mb algorithm. The first two experiments assess the
validity of \mb using a cross-validation design. Experiment 1 investigates the
accuracy of \mb whole hippocampus segmentation over a wide range of parameter
settings. This enables us to choose the parameter settings offering the best
performance for use in subsequent experiments. Experiment 2 tests hippocampal
subfield segmentation quality. The last two experiments assess the validity of
the \mb algorithm when applied to different diseases: Alzheimer's disease
(Experiment 3) and first episode schizophrenia patients (Experiment 4).

%The ADNI experiment includes comparison of volumes with other automated
%segmentation methods. Evaluation is done by correlating volumes b/c overlap
%metrics assume identical segmentation protocol.

@@

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Manual segmentations of the left and
right whole hippocampi are available. These labels have been generated using the
SNT tool from Medtronic Surgical Navigation Technologies, Louisville, CO (see
Supplementary Materials for detailed discussion of the manual segmentation
process used).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% ADNI1 X-Validation Dataset Demographics Table
%
<<ADNI-Xval-demographics, echo=F, results="asis",cache=FALSE>>=
adni_69_RIDs = c(295,413,619,685,729,782,938,954,1018,1155,907,981,222,324,448,
                 546,553,572,602,610,814,1341,675,681,731,1130,41,68,70,101,
                 249,293,316,414,698,1206,1222,1339,751,1030,3,5,10,16,22,53,
                 168,183,241,1205,328,1095,991,213,159,337,343,642,753,1109,
                 1217,29,56,1188,1293,149,382,431,1202)
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               adni_69_RIDs, method = "reverse", test=FALSE, overall=TRUE)

latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption="ADNI-1 cross-validation subset demographics", 
      title="tab:ADNI1-xvalidation-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Monte Carlo Cross-Validation (MCCV), also known as repeated random sub-sampling
cross-validation, consists of repeated rounds of validation conducted on a fixed
dataset \cite{Shao1993}. In each round, the dataset is randomly partitioned into
a training set and a validation set. The method to be validated is then given
the training data, and its output is compared with the validation set.

@@

A total of ten validation rounds are perfomed on each subject in the dataset,
over each combination of parameter settings. The parameter settings
we explore are: atlas library size (1-9), template library size (1-20),
registration method (\ants or \animal), and label fusion method (majority vote,
cross-correlation weighted majority vote, and normalized mutual information
weighted majority vote).  A total of $10 \times 69 \times 9 \times 20 \times 2
\times 3 = \num{7452000}$ validation rounds were conducted, resulting in a total
of $\num{1490400}$ segmentations analysed.

@@

The \animal algorithm carries out image registration is in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the whole
brain \citep{Collins}. In the second phase, nonlinear registration is completed
using the \animal algorithm \citep{Collins1995}: an iterative procedure that
estimates a 3D deformation field between two MR images. At first, large
deformations are estimated using blurred version of the input data. These larger
deformations are then input to subsequent steps where the fit is refined by
estimating smaller deformations on data blurred with a Gaussian kernel with a
smaller FWHM. The final transformation is a set of local translations defined on
a bed of equally spaced nodes that were estimated through the optimization of
the correlation coefficient.  For the purposes of this work we used the
regularization parameters optimized in Robbins et al. \citep{Robbins2004},
displayed in table \ref{tab:ANIMAL-params}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ANIMAL parameters                    
%                                     
\begin{table}[!tbp]
\begin{center}
\caption{\animal  registration parameters}
\begin{tabular}{l | c c c}
\hline 
Parameters        & Stage 1 & Stage 2 & Stage 3  \tabularnewline
\hline 
Model Blur (FWHM) &   8     &    8    &   4      \tabularnewline
Input Blur (FWHM) &   8     &    8    &   4      \tabularnewline
Iterations        &   30    &    30   &   10     \tabularnewline
Step              &   8x8x8 &  4x4x4  &   2x2x2  \tabularnewline
Sub-Lattice       &   6     &    6    &   6      \tabularnewline
Lattice Diameter  &24x24x24 &12x12x12 & 6x6x6    \tabularnewline
\hline
\end{tabular}
\end{center}
\label{tab:ANIMAL-params}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\ants is a diffeomorphic registration algorithm which provides great flexibility
over the choice of transformation model, objective function, and the consistency
of the final transformation. The transformation is estimated in a hierarchical
fashion where the MRI data is subsampled, allowing large deformations to be
estimated and successively refined at later hierarchical stages (where the data
is subsampled to a finer grid). The deformation field and the objective function
are regularized with a Gaussian kernel at each level of the hierarchy. The \ants
algorithm is freely available \url{http://www.picsl.upenn.edu/ANTS/}. We used an
implementation of the \ants algorithm compatible with the MINC data format,
mincANTS \url{https://github.com/vfonov/mincANTS}.

@@

The Dice similarity coefficient (DSC) assesses the agreement between two
segmentations. It is one of the most widely used measures of segmentation
performance, and we use it as the basis of comparison in this experiment.
Additionally, we report the Jaccard index, another commonly used similarity
measure:

@@

 \[\text{Jaccard (J)} = \frac{|A \cap B|}{|A \cup B|} = \frac{DSC}{(2-DSC)}\]

@@

The manual segmentations (SNT) provided as part of the ADNI dataset are used as 
the gold standard to compare with.  The segmentation accuracy reported is averaged 
over the ten validation rounds for each parameter setting.
